Forecasting short-term hourly Emergency Department arrivals

### 2022-01-12 paper/paper.Rmd
correct typo

---
title:  Forecasting short-term hourly Emergency Departement arrivals
author:
  - name: Bahmna Rostami Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliation: Cardiff Business School
    footnote: 12
  - name: Jethro Browell
    email: jethro.browell@glasgow.ac.uk
    affiliation: University of Glasgow
    footnote: 1
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Lancaster University
    footnote: 1
address:
  - code: Cardiff Business School
    address: Cardiff Business School, Cardiff University, UK
  - code: University of Glasgow
    address: School of Mathematics and Statistics, University of Glasgow, UK
  - code: Lancaster University
    address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
footnote:
  - code: 12
    text: "Corresponding Author"
  - code: 1
    text: "Equal contribution"

abstract: |
   An accurate forecast of Emergency Department (ED) arrivals by hour of the day is cirtical to meet patients's demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals. We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regresion, Prophet and a simple empirical distribution. We use Root Mean Squared Error (RMSE) to examin the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and  Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance or clinical desk services.

>>>>

---
title:  Forecasting short-term hourly Emergency Department arrivals
author:
  - name: Bahmna Rostami Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliation: Cardiff Business School
    footnote: 12
  - name: Jethro Browell
    email: jethro.browell@glasgow.ac.uk
    affiliation: University of Glasgow
    footnote: 1
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Lancaster University
    footnote: 1
address:
  - code: Cardiff Business School
    address: Cardiff Business School, Cardiff University, UK
  - code: University of Glasgow
    address: School of Mathematics and Statistics, University of Glasgow, UK
  - code: Lancaster University
    address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
footnote:
  - code: 12
    text: "Corresponding Author"
  - code: 1
    text: "Equal contribution"

abstract: |
   An accurate forecast of Emergency Department (ED) arrivals by hour of the day is critical to meet patients’ demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals. We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regression, Prophet, and a simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance, or clinical desk services.

### 2022-01-12 paper/paper.tex
correct typo

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Forecasting short-term hourly Emergency Departement arrivals},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Forecasting short-term hourly Emergency Departement arrivals}
    \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
   \ead{rostami-tabarb@cardiff.ac.uk} 
    \author[University of Glasgow]{Jethro Browell\corref{1}}
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[Lancaster University]{Ivan Svetunkov\corref{1}}
   \ead{i.svetunkov@lancaster.ac.uk} 
      \address[Cardiff Business School]{Cardiff Business School, Cardiff University, UK}
    \address[University of Glasgow]{School of Mathematics and Statistics, University of Glasgow, UK}
    \address[Lancaster University]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK}
      \cortext[12]{Corresponding Author}
    \cortext[1]{Equal contribution}

  \begin{abstract}
  An accurate forecast of Emergency Department (ED) arrivals by hour of the day is cirtical to meet patients's demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals. We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regresion, Prophet and a simple empirical distribution. We use Root Mean Squared Error (RMSE) to examin the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance or clinical desk services.
  \end{abstract}
   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
 \end{frontmatter}

@@

\ref{fig:time-accuracy} reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Forecasting short-term hourly Emergency Department arrivals},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Forecasting short-term hourly Emergency Department arrivals}
    \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
   \ead{rostami-tabarb@cardiff.ac.uk} 
    \author[University of Glasgow]{Jethro Browell\corref{1}}
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[Lancaster University]{Ivan Svetunkov\corref{1}}
   \ead{i.svetunkov@lancaster.ac.uk} 
      \address[Cardiff Business School]{Cardiff Business School, Cardiff University, UK}
    \address[University of Glasgow]{School of Mathematics and Statistics, University of Glasgow, UK}
    \address[Lancaster University]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK}
      \cortext[12]{Corresponding Author}
    \cortext[1]{Equal contribution}

  \begin{abstract}
  An accurate forecast of Emergency Department (ED) arrivals by hour of the day is critical to meet patients' demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals. We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regression, Prophet, and a simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance, or clinical desk services.
  \end{abstract}
   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
 \end{frontmatter}

@@

Figure \ref{fig:time-accuracy} reports the forecast accuracy of each approach versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours.
